Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-03-17 paperDistance/paper.tex
figures

\section{Why does this work?}

@@

We make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the Discussion below.
\begin{itemize}
\item We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.  In detail, this assumption is wrong, of
  course (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak; as long
  as there is support in the high signal-to-noise stars for the types
  of stars seen at low signal-to-noise, the results will not be
  strongly biased, but this assumption is deep and fundamental to this
  work.
\item We assume that we have large numbers of stars, large enough that
  an empirical-Bayes or maximum-marginalized likelihood is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case of a stellar type with only one (statistically distinct)
  exemplar in the data set, this assumption will be maximally wrong.
\item We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the [INSERT
    CORRECT TERMINOLOGY FROM GAIA HERE] parallax uncertainty estimates
  are correct. Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative
  (it leads to under-deconvolution).
\item We treat the \panstarrs-based three-dimensional dust maps (CITE
  SCHLAFLY) as correct in their median dust estimates, and their
  effects on color and magnitude. We iteratively update each star's
  dust correction as its inferred distance updates as we optimize the
  empirical-Bayes model. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for either
  color or absolute magnitude.
\item We assume that the \cmd\ can be represented with a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128, 512, and 2048 Gaussians in
  different experiments.
\item We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{itemize}

We would like to infer the true parallax $\varpi_{\true}$ for
each star given its parallax measurement $\varpi$ and associated
uncertainty $\sigma_{\varpi}$ from \tgas.
We assume (as \gaia\ suggests) that the likelihood is Gaussian
\begin{equation}
p(\varpi \given \varpi_{\true}, I) = \mathcal{N}(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \quad ,
\label{eq:gaialike}
\end{equation}
where $I$ is a placeholder containing our assumptions, including all
uncertainty estimates, and $\mathcal{N}(x \given \mu, \sigma^2)$ is a
Gaussian function for $x$ with mean $\mu$ and variance $\sigma^2$.
This is the probability of the data given the true parallax, but we
want to flip it around and infer the true parallax, or distance, to
the star given the measured parallax and it's uncertainty. We do this
using probability theory
\begin{equation}
p(\varpi_{true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{true}, \sigma^2_{\varpi}) \, p(\varpi_{true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.
The strength of our inference method here lies in how we generate our prior
pdf for the parallax $p(\varpi_{true})$.
Instead of assuming a functional form for the spatial distribution of
stars in the Milky Way (CITE TRI--CORYN), or relying on stellar
models, we build a noise-deconvolved estimate of the color--magnitude
diagram (\cmd) from the observed colors and absolute magnitudes of our
full set of stars.
This \cmd\ model becomes, for each star, a prior pdf for the distance
to that star.

[HOGG: NOW CONNECT THIS TO WHAT WE ACTUALLY DO WITH THE 2-d SPACE]

@@

\section{Data and results}

@@

@@

LMA: what are our results?
\subsection{Distances to M67}
As proof of concept we inferred the distances to stars in the open cluster M67. It is an open cluster in the constellation Cancer and is estimated to be 3-5 billion years old and 800-900 pc away. Previous work has estimated that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the signal to noise of the parallax measurement drops below 5. With \gaia\ parallaxes having a minimum uncertainty of about 0.3 mas, for stars with the most certain parallaxes, the SN reaches 5 at about 700 pc, right where this cluster lies. So it is a great test case for comparing parallax inferences with or without a prior. We made a window of 1 degree squared on the sky with the cluster M67 lying in the middle. The cluster is visibly obvious as a clustering of stars in a visualization of the \gaia\ data in 2D on the sky. With our converged dust values, and it's associated prior, we calculate the posterior of the true parallaxes to the stars within this window on the sky around M67. Figure XXX shows the comparison of the likelihoods and posteriors. The vertical lines bracket the parallaxes of, or distances to, the stars associated with the cluster. The likelihoods of the observed parallaxes are far broader and the cluster is not obvious at all. In the posteriors there is a sharp peak in the probability at the previously determined distance to the cluster, the vertical lines, for far more stars than the likelihoods would imply. Our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy. Too strong?

>>>>

\section{Why does this work?}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{lineExample.png}
\caption{Why does this work: Fitting a line}
\label{fig:line}
\end{figure}

@@

We make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the Discussion below.
\begin{itemize}
\item We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.  In detail, this assumption is wrong, of
  course (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak; as long
  as there is support in the high signal-to-noise stars for the types
  of stars seen at low signal-to-noise, the results will not be
  strongly biased, but this assumption is deep and fundamental to this
  work.
\item We assume that we have large numbers of stars, large enough that
  an empirical-Bayes or maximum-marginalized likelihood is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case of a stellar type with only one (statistically distinct)
  exemplar in the data set (e.g. a white dwarf), this assumption will be maximally wrong.
\item We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the [INSERT
    CORRECT TERMINOLOGY FROM GAIA HERE] parallax uncertainty estimates
  are correct. Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative
  (it leads to under-deconvolution).
\item We treat the \panstarrs-based three-dimensional dust maps (CITE
  SCHLAFLY) as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical-Bayes model, we iteratively update each star's
  dust correction as its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for either
  color or absolute magnitude.
\item We assume that the \cmd\ can be represented with a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128, 512, and 2048 Gaussians in
  different experiments.
\item We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact). Dust model dependent on stellar models?
\end{itemize}

Given these assumptions, we would like to infer the true parallax $\varpi_{true}$ to a star using the measured noisy parallax $\varpi$ and $\sigma_{\varpi}$ from \gaia. \gaia\ gives us the likelihood of the parallax of a star, given some unknown true parallax and uncertainties, which we assume to be a Gaussian
\begin{equation}
p(\varpi \given \varpi_{true}, \sigma_{\varpi},I) = \mathcal{N}(\varpi \given \varpi_{true}, \sigma_{\varpi})
\label{eq:gaialike}
\end{equation}

where $\varpi_{true}$ represents the true parallax, or distance, to the star, $\varpi$ is the observed parallax, and $\sigma_{\varpi}$ is the uncertainty in that measurement. There are other various prior assumptions (mostly about the noise properties which we will drop in our notation but is implicitly there and labeled $I$ below). $\mathcal{N}$ is the Gaussian function $\mathcal{N}(x \given \mu, \sigma)$ with mean $\mu$ and variance $\sigma$. This is the probability of the data given the true parallax, but we want to flip it around and infer the true parallax, or distance, to the star given the measured parallax and it's uncertainty. We do this using probability theory/bayes rule, and turn the likelihood of the data into the posterior probability of the true parallax

\begin{equation}
p(\varpi_{true} \given \varpi, \sigma^2_{\varpi}) = \frac{p(\varpi \given \varpi_{true}, \sigma^2_{\varpi}) \, p(\varpi_{true})}{Z},
\label{eq:bayes}
\end{equation}

where $p(\varpi_{true})$ is our prior knowledge of the parallax to the star, and $Z$ is a normalization constant to properly normalize the posterior. The strength in our inference method lies in how we generate our prior knowledge of the parallax to a star, $p(\varpi_{true})$. For our prior, instead of assuming the spatial distribution of stars in the Milky Way, or relying on stellar models, as previous work as shown, we build the \cmd\ from the observed colors and absolute magnitudes of our full set of stars. The absolute magnitude of the \cmd\ carries information about the distances, or parallaxes, to stars and the tight correlation of observed stars within this space dictates where each indivdually observed star can physically lie in this space. Toy model?

@@

\section{Data and results}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{data.png}
\caption{The observed \cmd: point estimates on the left and including error bars on the right.}
\label{fig:data}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance projected on the l,b galactic coordinates. This shows both the footprint of our analysis as well as the dust values we used. Each point represents a single star in our cross matched catalog. The large missing areas are due to the lack of dust values from \panstarrs\ since it only looked above -30 degrees in dec }
\label{fig:dust}
\end{figure}

@@

\subsection{Prior}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{priorNgaussComparison.png}
\caption{Prior: the XD generated mixture of gaussians we use as our prior}
\label{fig:prior}
\end{figure}

LMA: what are our results?
\subsection{Distances to M67}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67}
\label{fig:m67}
\end{figure}

As proof of concept we inferred the distances to stars in the open cluster M67. It is an open cluster in the constellation Cancer and is estimated to be 3-5 billion years old and 800-900 pc away. Previous work has estimated that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the signal to noise of the parallax measurement drops below 5. With \gaia\ parallaxes having a minimum uncertainty of about 0.3 mas, for stars with the most certain parallaxes, the SN reaches 5 at about 700 pc, right where this cluster lies. So it is a great test case for comparing parallax inferences with or without a prior. We made a window of 1 degree squared on the sky with the cluster M67 lying in the middle. The cluster is visibly obvious as a clustering of stars in a visualization of the \gaia\ data in 2D on the sky. With our converged dust values, and it's associated prior, we calculate the posterior of the true parallaxes to the stars within this window on the sky around M67. Figure XXX shows the comparison of the likelihoods and posteriors. The vertical lines bracket the parallaxes of, or distances to, the stars associated with the cluster. The likelihoods of the observed parallaxes are far broader and the cluster is not obvious at all. In the posteriors there is a sharp peak in the probability at the previously determined distance to the cluster, the vertical lines, for far more stars than the likelihoods would imply. Our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy. Too strong?

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{5 example posteriors showing the priors and likelihoods as well}
\label{fig:posterior}
\end{figure}
